Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-05 tagless/tagless-final.tex
Minor editorial changes

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless evaluator
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. 
\oleg{Jacques, do we use final and initial here correctly?}
Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.  \Citet{Gluck-jones-optimality}
explains deeper technical reasons that inevitably lead to these performance
penalties.

@@

Our contributions are
\begin{enumerate}
\item to evaluate in direct or continuation-passing styles, 
  compile, and partially evaluate a typed higher-order object language
   in a typed metalanguage with a simple type system. All these
   interpreters have no type tags and never get stuck;
\item a Hindley-Milner--typed partial evaluator with no 
  polymorphic lift operation and with
  the delayed binding-time analysis of functions . The partial
  evaluator needs no advanced type systems such as GADT or intentional
  type analysis;
\item to check statically for well-typed and closed object programs
      using the type system \oleg{typechecker?} of the metalanguage;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CPS;
\item a clear way to extend the object language with more features
  such as state;
\item and an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out to be harder than expected.
\end{enumerate}

@@

In contrast to \S\ref{tagproblem}, we encode 
binding in object programs using higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices. This makes the encoding convenient and
also ensures that our object programs are closed.

@@

The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we will frequently show implementations of
only one method of each group and elide the rest, to save space. The
accompanying code has the complete code.

Because the types in the |Symantics| class reflect the object language's type
system, this representation admits only well-typed object terms for evaluation.
\oleg{replace the above with the following?}
The comparison of |Symantics| with Figure~\ref{fig:object} clearly
shows that every typed, closed object term can be represented in the
metalanguage. Moreover,
\begin{prop}
If an object term has the object type |t|, its
representation in the metalanguage has the type 
|forall repr. Symantics repr => repr t|
\oleg{or, has the type repr t for some repr. Meta-language embeds all
  object types.}
\end{prop}
If we make a mistake somewhere, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

@@

To embed the same object language in (Meta)OCaml, we replace the type
class |Symantics| and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until \S\ref{PE}.
Two differences between this signature and the |Symantics| class in
Haskell are: the additional type parameter |c|, an
environment classifier \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| due to OCaml's being a call-by-value
language.

The functor below encodes our running examples |test1| and $\mathit{power}$.
\begin{code}
module EX(S: Symantics) = struct
  open S
  let test1 () =
    app (lam (fun x -> x)) (bool true)
  let testpowfix () = 
    lam (fun x -> fix (fun self -> lam (fun n ->
    if_ (leq n (int 0)) (fun () -> int 1)
        (fun () -> mul x
          (app self (add n (int (-1))))))))
  let testpowfix7 = 
     lam (fun x -> app (app (testpowfix ()) x)
                       (int 7))
end;;
\end{code}
The dummy argument to |test1| and |testpowfix| is an artefact of
MetaOCaml, again related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction then dictates that
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R|
module.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
That is, the proposition below follows immediately from the soundness of
the metalanguage's type system.
\begin{prop}
If the term |e| encoded in the metalanguage has type~|t|,
then evaluating~|e| in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~|t|.
\end{prop}
In general, 
\begin{prop}
  If the definitional interpreter implementing |Symantics| 
  in the soundly typed metalanguage never gets stuck,
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  the interpreter.
\end{prop}

For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors. The following is slightly abbreviated code (see the
accompanying source code for the complete definition).
\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let lam f        = f 0 + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = f 0 + 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
end;;
\end{code}
Now the expression
\begin{code}
let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms like
\begin{code}
app (fix (fun self -> self)) (int 1)
\end{code}

@@

Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts, the last of which finally uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension to the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly speaking, a partial evaluator interprets each object term either
to a static (present-stage) term, as produced by the evaluator~|R|, or
to a dynamic (future-stage) term, as produced by the compiler~|C|.  The
interpreter computes as much as possible statically.  Thus one might try
to define partial evaluation as the Haskell data type
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrInt :: P0 Int -> C Int
abstrInt (S0 r) = int (unR r)
abstrInt (E0 c) = c

abstrBool :: P0 Bool -> C Bool
abstrBool (S0 r) = bool (unR r)
abstrBool (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x) (int x)
  bool x = S0 (bool x) (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add e1 e2 = E0 (add (abstrInt e1)
                      (abstrInt e2))
\end{code}
Integer and boolean literals are obviously immediate, present-stage
values. Addition yields a static term if and only if both operands
are static. If so, we use the evaluator~|R| to add the operands.
If not, we extract dynamic terms from the operands and add them
using the compiler~|C|.

@@

When we come to functions, however, we stumble.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
}|P1 (a -> a)|, consists of a dynamic part (type |C1 (a -> a)|) and
maybe a static part (type |R1 (a -> a)|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  Even the same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

@@

This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in |app|
above: the type |P t| has four constructors, of which pattern\hyp
matching in |app| uses only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
Indeed the metalanguage implementation could reason thus, but it may not; GHC for one
issues warnings.  Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but are total in reality. The goal is to make this totality
\emph{syntactically} apparent.

@@

Unlike the GADT approach in~\S\ref{S:PE-GADT}, all pattern\hyp matching
in~|P| is \emph{syntactically} exhaustive, so it is patent to the metalanguage
implementation that |P| is total.  Further, all pattern\hyp matching occurs
during partial evaluation, only to check if a value is known statically,
never what type it has.  In other words, our partial evaluator tags
phases (with |Some| and |None|) but not object types.

@@

We show that our |Symantics| and generally our approach accomodates
a number of variants without any (new) difficulties.  In particular,
we show how a a call-by-name CPS interpreter, a call-by-value CPS
program transformation, as well as how to deal with imperative features.
Finally, we comment on the requirements on the type system to support 
our approach.

@@

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given already by 
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is continuation-passing style (CPS), which they
specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting \oleg{mention the quote from
Reynolds that typed setting would be nice?}\jacques{I think so, but from
which paper? The one ReynoldsPlotkin paper is typed}\ccshan{I think Oleg
meant ``It should also be possible to define languages with a highly
 refined syntactic type structure. Ideally, such a treatment should be
 metacircular, in
 the sense that the type structure used in the defined language should be adequate for the
 defining language'' \citep{reynolds-definitional}, but perhaps Jacques
 meant another paper by ``ReynoldsPlotkin''?}. Specifically, we
demonstrate CBN evaluation for our object language in OCaml, which is
call-by-value. We do so by building a CBN CPS interpreter for the object
language.

The CBN CPS interpretation was introduced by \citet{PlotkinCBN}. As in
the more familiar CBV CPS, the interpretation of a value is a function
accepting a continuation argument |k| and yielding the answer
returned by that continuation.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = 
  fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1+v2)))
\end{code}
In both cases, the returned value has type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

@@

Because |RCN| has the signature |Symantics|, we can instantiate our previous
examples with it, and all works as expected.  More interesting
is the following example.
\begin{code}
module EXS(S: Symantics) = struct
 open S
 let diverg () = 
   app (lam (fun x -> int 1)) 
       (app (fix (fun self -> self)) (int 2))
end
\end{code}

If we interpret |EXS| with the |R| interpreter of
\S\ref{S:interpreter-RL} realized in OCaml
\begin{code}
let module M = EXS(R) in M.diverg ();;
\end{code}
we never get any result because the above diverges. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

@@

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = 
    fun k -> k (fun e -> e.ko 
         (fun v -> f {ko = fun k -> k v}))}
end;;
\end{code}
Now when we apply an abstraction to an argument |e|, we always force
the evaluation of that argument before proceeding. This approach is in
line with \citet{reynolds-relation} (albeit in a typed setting). The interpreter RCV
is useful for achieving CBV evaluation of the object language
no matter whether the metalanguage is CBV or CBN.

@@

To save space, we show a simplified and less principled approach that
lets us use the module |CPST| as it is. Because the module does not
have the signature Symantics, we cannot apply the |EX| functor to it.
But we can write the tests nevertheless:
\begin{code}
module T = struct
 module M = CPST(C)
 open M
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = ... (* the same as before *)
 let testpowfix7 = (* the same as before *)
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
That is, we instantiate |CPST| with the desired base intrepreter (the
compiler |C| in the case above) so that we can use the result |M| to
interpret object level terms. The sample terms are \emph{exactly}
as we used before. We just copied them over. That copying step is the
price we pay for the simplified treatment. We shall also encounter this
copying when discussing the self-interpreter in the next section. We
will see then that the copying step is not frivolous, but represents
a profound instance of 
``plugging the hole'', which is one of the many faces of polymorphism.

@@

We must formulate some propositions: a typed term makes progress in
any interpreter. The reason it is typed in any interpreter:
|forall r. Symantics r => r tau| directly says that in every model the
term is typed.

@@

The attempt to encode self-interpreter brings the |*->*| polymorphism
the the forefront, as we shall now see.

>>>>

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless evaluator
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. 
\oleg{Jacques, do we use final and initial here correctly?}
Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error (and is in fact
total).  Because this safety is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.  \Citet{Gluck-jones-optimality}
explains deeper technical reasons that inevitably lead to these performance
penalties.

@@

Our contributions are as follows.
\begin{enumerate}
\item We build interpreters that evaluate (in direct or continuation\hyp passing styles),
  compile, and partially evaluate a typed higher-order object language
   in a typed metalanguage.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.
\item We use the type system of the metalanguage
    to check statically for well-typed and closed object programs.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CPS transformation.
\item We point a clear way to extend the object language with more features
  such as state.
\item We describe an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out harder than expected.
\end{enumerate}

@@

In contrast to \S\ref{intro}, we henceforth encode 
binding in object programs using higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices. This makes the encoding convenient and
also ensures that our object programs are closed.

@@

The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we will frequently show implementations of
only one method of each group and elide the rest, to save space. The
accompanying code has the complete implementations.

Comparing |Symantics| with Figure~\ref{fig:object}
shows how to represent every typed, closed object term in the
metalanguage. Moreover, the representation preserves types, in the following
sense.
\begin{prop}
If an object term has the object type |t|, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr t|.
\end{prop}
Conversely, the type system of the metalanguage checks statically that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

@@

To embed the same object language in (Meta)OCaml, we replace the type
class |Symantics| and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.
Two differences between this signature and the |Symantics| class in
Haskell are: the additional type parameter |c|, an
environment classifier \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| due to OCaml's being a call-by-value
language.

The functor below encodes our running examples |test1| and $\mathit{power}$.
\begin{code}
module EX(S: Symantics) = struct
  open S
  let test1 () =
    app (lam (fun x -> x)) (bool true)
  let testpowfix () = 
    lam (fun x -> fix (fun self -> lam (fun n ->
    if_ (leq n (int 0)) (fun () -> int 1)
        (fun () -> mul x
          (app self (add n (int (-1))))))))
  let testpowfix7 = 
     lam (fun x -> app (app (testpowfix ()) x)
                       (int 7))
end
\end{code}
The dummy argument to |test1| and |testpowfix| is an artefact of
MetaOCaml, again related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction then dictates that
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R|
module.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{prop}
If the term |e| encoded in the metalanguage has type~|t|,
then evaluating~|e| in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~|t|.
\end{prop}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{prop}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{prop}
These propositions follow immediately from the soundness of the
metalanguage's type system.

For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors. The following is slightly abbreviated code (see the
accompanying source code for the complete definition).
\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let lam f        = f 0 + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = f 0 + 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
end
\end{code}
Now the expression
\begin{code}
let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms like
\begin{code}
app (fix (fun self -> self)) (int 1)
\end{code}

@@

Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension to the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly speaking, a partial evaluator interprets each object term either
to a static (present-stage) term, as produced by the evaluator~|R|, or
to a dynamic (future-stage) term, as produced by the compiler~|C|.  To
distinguish between static and dynamic terms, one might try
to define partial evaluation as the Haskell data type
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrInt :: P0 Int -> C Int
abstrInt (S0 r) = int (unR r)
abstrInt (E0 c) = c

abstrBool :: P0 Bool -> C Bool
abstrBool (S0 r) = bool (unR r)
abstrBool (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add e1 e2 = E0 (add (abstrInt e1)
                      (abstrInt e2))
\end{code}
Integer and boolean literals are obviously immediate, present-stage
values. Addition yields a static term if and only if both operands
are static. If so, we use the evaluator~|R| to add the operands.
If not, we extract dynamic terms from the operands and add them
using the compiler~|C|.
Thus we use the overloaded functions |int| and |bool| from the |Symantics|
instances for |R| and~|C| to define those for~|P0|.

@@

When we come to functions, however, we stumble.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
}|P1 (a -> a)|, consists of a dynamic part (type |C (a -> a)|) and
maybe a static part (type |R (a -> a)|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  Even the same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

@@

This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in |app|
above: the type |P t| has four constructors, of which pattern\hyp
matching in |app| uses only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
Indeed the metalanguage implementation could reason thus, but it may not; GHC for one
issues warnings.  Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but never fail in reality. The goal is to make this safety
\emph{syntactically} apparent.

@@

Unlike the GADT approach in~\S\ref{S:PE-GADT}, all pattern\hyp matching
in~|P| is \emph{syntactically} exhaustive, so it is patent to the metalanguage
implementation that |P| never gets stuck.  Further, all pattern\hyp matching occurs
during partial evaluation, only to check if a value is known statically,
never what type it has.  In other words, our partial evaluator tags
phases (with |Some| and |None|) but not object types.

@@

We show that our |Symantics| and generally our approach accomodates
a number of variants without any (new) difficulties.  In particular,
we show how a call-by-name CPS interpreter, a call-by-value CPS
program transformation, as well as how to deal with imperative features.
Finally, we comment on the requirements on the type system to support 
our approach.

@@

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given already by 
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes thanks to CPS, which they
specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting \oleg{mention the quote from
Reynolds that typed setting would be nice?}\jacques{I think so, but from
which paper? The one ReynoldsPlotkin paper is typed}\ccshan{I think Oleg
meant ``It should also be possible to define languages with a highly
 refined syntactic type structure. Ideally, such a treatment should be
 metacircular, in
 the sense that the type structure used in the defined language should be adequate for the
 defining language'' \citep{reynolds-definitional}, but perhaps Jacques
 meant another paper by ``ReynoldsPlotkin''?}. Specifically, we
demonstrate CBN evaluation for our object language in OCaml, which is
call-by-value. We do so by building a CBN CPS interpreter for the object
language.

The CBN CPS interpretation was introduced by \citet{PlotkinCBN}. As in
CBV CPS, the interpretation of an object term is a function
mapping a continuation~|k| to the answer
returned by~|k|.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = fun k ->
  e1 (fun v1 -> e2 (fun v2 -> k (v1 + v2)))
\end{code}
In both cases, the returned value has type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

@@

Because |RCN| has the signature |Symantics|, we can instantiate our previous
examples with it, and all works as expected.  More interesting
is the example $(\fun{x}1)\bigl((\fix{f}f)\mathinner2\bigr)$, which terminates
under CBN but not CBV\@.
\begin{code}
module EXS(S: Symantics) = struct
 open S
 let diverg () = 
   app (lam (fun x -> int 1)) 
       (app (fix (fun f -> f)) (int 2))
end
\end{code}

If we interpret |EXS| with the |R| interpreter of
\S\ref{S:interpreter-RL} realized in OCaml
\begin{code}
let module M = EXS(R) in M.diverg ()
\end{code}
we never get any result because the above diverges. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

@@

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = 
    fun k -> k (fun e -> e.ko 
         (fun v -> f {ko = fun k -> k v}))}
end
\end{code}
Now when we apply an abstraction to an argument |e|, we always force
the evaluation of that argument before proceeding. This approach is in
line with \citet{reynolds-relation} (albeit in a typed setting). The interpreter RCV
is useful for achieving CBV evaluation of the object language
no matter whether the metalanguage is CBV or CBN.

@@

To save space, we show a simplified and less principled approach that
lets us use the module |CPST| as it is. Because the module does not
have the signature Symantics, we cannot apply the |EX| functor to it.
But we can write the tests nevertheless:
\begin{code}
module T = struct
 module M = CPST(C)
 open M
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = ... (* the same as before *)
 let testpowfix7 = (* the same as before *)
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
That is, we instantiate |CPST| with the desired base intrepreter (the
compiler |C| in the case above) so that we can use the result |M| to
interpret object level terms. The sample terms are \emph{exactly}
as we used before. We just copied them over. That copying step is the
price we pay for the simplified treatment. We shall also encounter this
copying when discussing the self-interpreter in the next section. We
will see then that the copying step is not frivolous, but represents
a profound instance of 
``plugging the hole'', which is one of the many faces of polymorphism.

@@

@@

The attempt to encode self-interpreter brings the |*->*| polymorphism
the forefront, as we shall now see.
